Gradual Soundness: Lessons from Static Python

### 2022-01-06 pj-7.1/paper.tex
pj: sec 3 soundness argument

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
%% - what are the idiomatic base SP types ... int or Int ?
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\Gamma}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\code{c}}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\spsubt}{\mathrel{<:}}
\newcommand{\spsubteq}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[1]{\spapp{\code{chkdict}}{#1}}
\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
\newcommand{\spdictref}[2]{#1[#2]}
\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
\newcommand{\spobjref}[2]{#1.#2}
\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}

@@

@@

To validate the design of \SP{}, we have developed a mechanized model using PLT
Redex~\cite{RYR}.
The model encompasses two soundness claims:
%% addresses two soundness concerns?
\begin{enumerate}
  \item
    \SP{}'s overall approach to gradual interactions is sound; namely, that interactions
    with untyped code do not undermine the validity of types.
  \item
    The typed half of \SP{} is a sound type system for Python.
\end{enumerate}
This section explores the first claim in depth by way of a tiny formal model
that illustrates the \SP{} gradual interaction strategy.
The second claim is justified by our Redex development;
in particular, by its test suite.
This section describes the Redex development in broad strokes.

FILL pick names to avoid confusion between the Redex model and the tiny paper model?

@@

\subsection{Surface Syntax and Types}

\begin{figure}[t]
  \begin{langarray}
    \spprog & \defeq &
      \spstmt \langmid \spstmt,~\spprog
  \\
    \spstmt & \defeq &
      \spvardef{\spx}{\sptype}{\spexpr} \langmid
  \\ & &
      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
  \\ & &
      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr}}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr}}
  \\
    \spexpr & \defeq &
      \spx \langmid \spnone \langmid \spobject \langmid \spint \langmid \spbool \langmid \spfloat \langmid
      \sppydict{\spx: \spexpr, \ldots} \langmid
      \spchkdict{\sppydict{\spx: \spexpr, \ldots}} \langmid
  \\ & &
      \spapp{\spx}{\spexpr} \langmid
      \spdictref{\spexpr}{\spexpr} \langmid
      \spdictset{\spexpr}{\spexpr}{\spexpr} \langmid
      \spobjref{\spc}{\spx} \langmid
      \spobjset{\spc}{\spx}{\spexpr} \langmid
      \spobjapp{\spc}{\spf}{\spexpr}
  \\
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptclass \langmid
      \sptobject \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptfloat \langmid
  \\ & &
      \sptpydict{\sptype}{\sptype} \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptunion{\sptype, \ldots}
  \\
    \sptenv & \defeq &
      \sptenvnil \langmid
      \sptvardef{\spx}{\sptype},~\sptenv \langmid
      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype}}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype}},~\sptenv
  \\
    \spx, \spf, \spc & \defeq & \mbox{variable names}
  \end{langarray}

%%  \begin{center}\parbox{0.8\columnwidth}{
  \bigskip
    \mbox{Abbreviation: $\sptoptional{\sptype} \defeq \sptunion{\sptnone, \sptype}$}
%%  }\end{center}

@@

%% TODO assume well-formed / normal-form
%% - no circular field defs (x:t = x)
%% - unions are tidy ... Opt[Dyn] = Dyn
%% kk

%% TODO surface typing judgment omitted
%% - straightforward
%% - eval typing far more interesting
%% - for now, main goal is to narrow unions ... that's kinda boring / not worth effort reading & writing

The \SP{} surface syntax is the same as Python 3.8.
Programmers may write type annotations using the PEP 484 syntax;
unannotated code gets the dynamic type by default.
The available static types provide a nominal type system with first-order functions.
These surface types are meant to be descriptive.
By design, they are not fully sound.

\Cref{f:surface-types} illustrates the key aspects of the \SP{} syntax and types
using a tiny grammar.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require full type annotations.
Functions must have one positional argument.
Classes must declare one parent (perhaps $\spobject$), one field, and one method.
Expressions describe values and simple computations.
The basic values consist of the following: the none value, the top object, integers,
booleans (technically the integers \code{0} and \code{1}), and floating point numbers.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
and method calls.
Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
There is no function type; the only way to define a higher-order function is to use the dynamic type.

Relative to \SP{} and our Redex model, the formalization in \cref{f:surface-types} omits
many details of Python including imports, conditionals, and exception handlers.
These details are crucial in the context of the Redex model, which tests whether \SP{}
soundly approximates Python.
They are less important here, where our focus is on type boundaries.
None of the omitted features give substantially new ways for typed code to interact with dynamic code.

\subsection{Evaluation Types, Materialization, Casts}

\begin{figure}[t]
  \fbox{$\sptenv \wtprog \spprog \dashv \sptenv$}
  \begin{mathpar}
    \inferrule*{
      \sptenv_0 \wtprog \spstmt_0 \dashv \sptenv_1
      \\\\
      \sptenv_1 \wtprog \spprog_0 \dashv \sptenv_2
    }{
      \sptenv_0 \wtprog \spstmt_0, \spprog_0 \dashv \sptenv_2
    }

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }{
      \sptenv_0 \wtprog \spvardef{\spx_0}{\sptype_0}{\spexpr_0} \dashv \sptvardef{\spx_0}{\sptype_0},~\sptenv_0
    }

    \inferrule*{
      \sptenv_1 = \sptfundef{\spf_0}{\sptype_0}{\sptype_1},~\sptenv_0
      \\\\
      \sptenv_1 \wtexpr \spexpr_0 : \sptype_1
    }{
      \sptenv_0 \wtprog \spfundef{\spf_0}{\spann{\spx_0}{\sptype_0}}{\sptype_1}{\spexpr_0} \dashv \sptenv_1
    }

    \inferrule*{
      \sptenv_1 = \sptclassdef{\spc_0}{\spc_1}{\sptvardef{\spx_0}{\sptype_0}}{\sptfundef{\spf_1}{\spc_0, \sptype_1}{\sptype_2}}, \sptenv_0
      \\\\
      %% TODO are field overrides exact?
      \mbox{if $\spann{\spx_0}{\sptype_0'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\sptype_0 = \sptype_0'$}
      \\\\
      %% TODO can T override dyn?
      \mbox{if $\sptfundef{\spf_1}{\spc_2, \sptype_1'}{\sptype_2'} \in \spenvapp{\sptenv_1}{\spc_1}$ then $\sptype_1' \spsubteq \sptype_1$ and $\sptype_2 \spsubteq \sptype_2'$}
      \\\\
      \sptenv_1 \wtexpr \spvardef{\spexpr_0} : \sptype_0
      \\
      \spann{\spself}{\spc_0}, \spann{\spx_1}{\sptype_1}, \sptenv_1 \wtexpr \spexpr_1 : \sptype_2
    }{
      \sptenv_0 \wtprog \spclassdef{\spc_0}{\spc_1}{\spvardef{\spx_0}{\sptype_0}{\spexpr_0}}{\spfundef{\spf_1}{\spself, \spann{\spx_1}{\sptype_1}}{\sptype_2}{\spexpr_1}} \dashv \sptenv_1
    }
  \end{mathpar}

  \fbox{$\sptenv \wtexpr \spexpr : \sptype$}~selected rules
  \begin{mathpar}
    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_1
      \\
      \sptype_1 \spcompat \sptype_0
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_1
      \\
      \sptype_1 \spmatr \sptype_0
    }{
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }

    \inferrule*{
      \sptvardef{\spf_0}{\sptdyn} \in \sptenv_0
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
    }{
      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \sptdyn
    }

    \inferrule*{
      \sptfundef{\spf_0}{\sptype_0}{\sptype_1} \in \sptenv_0
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \sptype_1
    }

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
    }{
      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
    }

    \inferrule*{
      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\sptype_0}{\sptype_1}
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptnone
    }

    \inferrule*{
      \spann{\spx_0}{\sptype_0} \in \spenvapp{\sptenv_0}{\spc_0}
    }{
      \sptenv_0 \wtexpr \spobjref{\spc_0}{\spx_0} : \sptype_0
    }

    \inferrule*{
      \spann{\spx_0}{\sptype_0} \in \spenvapp{\sptenv_0}{\spc_0}
      \\\\
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spobjset{\spc_0}{\spx_0}{\spexpr_0} : \sptnone
    }

    \inferrule*{
      \sptfundef{\spf_0}{\spc_0, \sptype_0}{\sptype_1} \in \spenvapp{\sptenv_0}{\spc_0}
      \\\\
      \sptenv_0 \wtexpr \spexpr_0 : \sptype_0
    }{
      \sptenv_0 \wtexpr \spobjapp{\spc_0}{\spf_0}{\spexpr_0} : \sptype_1
    }

>>>>

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
%% - what are the idiomatic base SP types ... int or Int ?
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%% - use ChkDict not ChkList ... avoid confusion
%% - is \sqle _consistent_ or _compatible_ subtyping?

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\typefont{Env}}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\trule}[1]{\textsc{#1}}

\newcommand{\spx}{\code{x}}
\newcommand{\spf}{\code{f}}
\newcommand{\spc}{\code{c}}
\newcommand{\spprog}{\codefont{prog}}
\newcommand{\spstmt}{\codefont{stmt}}
\newcommand{\spexpr}{\codefont{expr}}
\newcommand{\spvalue}{\codefont{val}}
\newcommand{\vdashsub}[1]{\vdash_{#1}}
\newcommand{\wtprog}{\vdashsub{\mathbf{P}}}
\newcommand{\wtexpr}{\vdashsub{\mathbf{E}}}
\newcommand{\spsubt}{\mathrel{<:}}
\newcommand{\spsubteq}{\mathrel{\leq:}}
\newcommand{\spcompat}{\sqsubseteq}
\newcommand{\spmatr}{\prec}
\newcommand{\spvardef}[3]{\sptvardef{#1}{#2} = #3}
\newcommand{\spfundef}[4]{\mathrm{def}~\spapp{#1}{#2}\code{ -> }#3: #4}
\newcommand{\spclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\spself}{\code{self}}
\newcommand{\spobject}{\code{object}}
\newcommand{\spnone}{\code{none}}
\newcommand{\spint}{\codefont{int}}
\newcommand{\spbool}{\codefont{bool}}
\newcommand{\spfloat}{\codefont{float}}
\newcommand{\sppydict}[1]{\code{\{}#1\code{\}}}
\newcommand{\spchkdict}[3]{\spapp{\paramtype{\code{chkdict}}{#1, #2}}{#3}}
\newcommand{\spenvapp}[2]{\spapp{#1}{#2}}
\newcommand{\spdictref}[2]{#1[#2]}
\newcommand{\spdictset}[3]{\spdictref{#1}{#2} = #3}
\newcommand{\spobjref}[2]{#1.#2}
\newcommand{\spobjset}[3]{\spobjref{#1}{#2} = #3}
\newcommand{\spobjapp}[3]{\spobjref{#1}{\spapp{#2}{#3}}}

@@

%% 2022-01-05 Carl:
%% "All the data I’ve been sharing is from a single codebase, the primary IG web
%% server monolith. The existence of JITted but not SP modules is mostly just due
%% to the fact that opting a function into the JIT is semantically transparent and
%% (unless it reveals a latent bug in the JIT) requires no code changes, so it’s
%% quite easy to do (the only tradeoff pushing us away from JITting everything is
%% just how much CPU we spend JIT-compiling and how much memory we spend storing
%% JIT-compiled code), whereas opting a module into SP requires more work so the
%% bottleneck is available developer time to do conversions."

@@

To validate the design of \SP{}, we have developed a mechanized model using PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model encompasses two soundness claims:
%% addresses two soundness concerns?
\begin{enumerate}
  \item
    \SP{}'s overall approach to gradual interactions is sound; namely, that interactions
    with untyped code do not undermine the validity of types.
  \item
    The typed half of \SP{} is a sound type system for Python.
\end{enumerate}
This section explores the first claim in depth by way of a tiny formal model
that illustrates the \SP{} gradual interaction strategy.
The second claim is justified by our Redex development;
in particular, by its test suite.
This section describes the Redex development in broad strokes.

@@

\subsection{Surface Syntax and Types}
%% TODO assume well-formed; no circular defs (x:t = x)

\begin{figure}[t]
  \begin{langarray}
    \spprog & \defeq &
      \spstmt \langmid \spstmt,~\spprog
  \\
    \spstmt & \defeq &
      \spvardef{\spx}{\sptype}{\spexpr} \langmid
  \\ & &
      \spfundef{\spf}{\spann{\spx}{\sptype}}{\sptype}{\spexpr} \langmid
  \\ & &
      \spclassdef{\spc}{\spc}{\spvardef{\spx}{\sptype}{\spexpr}}{\spfundef{\spf}{\spself, \spann{\spx}{\sptype}}{\sptype}{\spexpr}}
  \\
    \spexpr & \defeq &
      \spx \langmid \spnone \langmid \spint \langmid \spbool \langmid \spfloat \langmid \spobject \langmid
      \sppydict{\spx: \spexpr, \ldots} \langmid
      \spchkdict{\sptype}{\sptype}{\sppydict{\spx: \spexpr, \ldots}} \langmid
  \\ & &
      \spapp{\spx}{\spexpr} \langmid
      \spdictref{\spexpr}{\spexpr} \langmid
      \spdictset{\spexpr}{\spexpr}{\spexpr} \langmid
      \spobjref{\spc}{\spx} \langmid
      \spobjset{\spc}{\spx}{\spexpr} \langmid
      \spobjapp{\spc}{\spf}{\spexpr}
  \\
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptfloat \langmid
      \sptobject \langmid
      \sptclass \langmid
  \\ & &
      \sptpydict{\sptype}{\sptype} \langmid
      \sptchkdict{\sptype}{\sptype} \langmid
      \sptunion{\sptype, \ldots}
  \\
    \sptenv & \defeq &
      \sptenvnil \langmid
      \sptvardef{\spx}{\sptype},~\sptenv \langmid
      \sptfundef{\spf}{\sptype}{\sptype},~\sptenv \langmid
      \sptclassdef{\spc}{\spc}{\sptvardef{\spx}{\sptype}}{\sptfundef{\spf}{\sptclass, \sptype}{\sptype}},~\sptenv
  \\
    \spx, \spf, \spc & \defeq & \mbox{variable names}
  \end{langarray}

  \bigskip
  \mbox{Abbreviation: $\sptoptional{\sptype} \defeq \sptunion{\sptnone, \sptype}$}

@@

The \SP{} surface syntax is the same as Python 3.8.
Programmers may write type annotations using the PEP 484 syntax;
unannotated code gets the dynamic type by default.
The available static types provide a nominal type system with first-order functions.
These surface types are meant to be descriptive.
By design, they are not fully sound.
For this reason, we omit the surface typing judgment and merely present their
declarative syntax.
%% TODO awkward

\Cref{f:surface-types} illustrates the key aspects of the \SP{} syntax and types
using a tiny grammar.
A program is a sequence of statements;
a statement defines a variable, function, or class.
These definitions may only appear on the top level
and they all require full type annotations.
Functions must have one positional argument.
Classes must declare one parent (perhaps $\spobject$), one field, and one method.
Expressions describe values and simple computations.
The basic values consist of the following: the none value, integers,
booleans (which are the integers \code{0} and \code{1}), floating point numbers, and the top object.
There are two data structures: dictionaries and checked dictionaries~(\cref{s:checked-type}).
The remaining forms express function calls, dictionary reads and writes, object field reads and writes,
and method calls.
Types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We assume that all unions are written a concise normal form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be flattened and normalized to $\sptdyn$.
There is no function type; the only way to define a higher-order function is to use the dynamic type.

%% FILL generally, no way to express higher-order cast ... either trivial C=C or impossible C=D

Relative to \SP{} and our Redex model, the formalization in \cref{f:surface-types} omits
many details of Python including class variables, imports, conditionals, and exception handlers.
These details are crucial in the context of the Redex model, which tests whether \SP{}
soundly approximates Python.
They are less important here, where our focus is on type boundaries.
None of the omitted features give substantially new ways for typed code to interact with dynamic code.

\subsection{Evaluation Types, Casts, and Typing Judgment}

\begin{figure}[t]
  \begin{langarray}
    \spteval & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptint \langmid
      \sptbool \langmid
      \sptfloat \langmid
      \sptobject \langmid
      \sptclass \langmid
    \\ & &
      \sptrawpydict \langmid
      \sptchkdict{\spteval}{\spteval} \langmid
      \sptoptional{\spteval}
  \end{langarray}

  \bigskip
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptrawpydict & \mbox{if $\sptype_0 = \sptpydict{\sptype}{\sptype}$}
    \\
      \sptchkdict{\mftypeF{\sptype_1}}{\mftypeF{\sptype_2}} & \mbox{if $\sptype_0 = \sptchkdict{\sptype_1}{\sptype_2}$}
    \\
      \sptoptional{\mftypeF{\sptype_1}} & \mbox{if $\sptype_0 = \sptoptional{\sptype_1}$}
    \\
      \sptdyn & \mbox{if $\sptype_0 = \sptunion{\sptype, \ldots}$ and $\sptype_0 \neq \sptoptional{\sptype}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

  \caption{Evaluation Types, Surface-to-Evaluation Mapping}
  \label{f:surface-to-eval-types}
\end{figure}

Evaluation type $\spteval$ are a subset of the surface types.
These are the types that \SP{} promises to soundly enforce.
\Cref{f:surface-to-eval-types} presents the syntax of evaluation types
and a retraction $\mftypeF{\cdot}$ from surface types to evaluation types.
As the retraction makes clear, the evaluation types make two main changes:
\begin{enumerate}
  \item
    The parameterized type for Python dictionaries
    $\sptpydict{\sptype_0}{\sptype_1}$ gets replaced with
    a raw type $\sptrawpydict$.
    The raw type behaves the same as $\sptpydict{\sptdyn}{\sptdyn}$ would.
  \item
    Unions get replaced with the dynamic type, except for the special case
    of unions with none ($\sptoptional{\sptype_0}$).
\end{enumerate}

The \SP{} team chose these particular evaluation types because they can all
be enforced fully and efficiently with run-time casts.

\subsubsection{Casts}

\Cref{t:cast} describes the casts in detail.
Most call for a simple tag check, i.e., a Python \code{isinstance} test.
The sole exception is optional types, which require a tag check and a test for the none value.
Even checked dictionaries rely on tag checks.
A checked dict type with keys $\sptype_0$ and values $\sptype_1$ accepts only
checked dict values that were initialized with exactly the same key and value type.
As noted in~\cref{s:type-dynamic}, the exact-match rules applies even when $\sptype_0$
or $\sptype_1$ is the dynamic type.

There are two key observations to make about casts.
First, all allowed casts can run in constant time.
No cast requires traversing a data structure.
Similarly, no cast allocates a wrapper to check higher-order behaviors in a delayed fashion.
This latter property means that blame-tracking is trivial;
casts succeed or fail immediately.

Second, the strict cast for checked dictionaries makes type compatibility simple in \SP{}.
Values that are partly-dynamic are incompatible with static types.
For example, suppose that a function expects a list of numbers.
In a typical gradual language, a list of dynamic values would be a valid input to the function.
In \SP{}, a list of dynamic values is not a valid input.
Consequently, \SP{} can use a simple ahead-of-time strategy to enforce soundness.
Any position where the dynamic type meets a static type needs a cast.
Other meetings between types can be resolved statically.

\begin{table}[t]
  \caption{How to Enforce the Evaluation Types}
  \label{t:cast}

  \(\begin{array}{l@{\quad}l}
    \mbox{Eval. Type $\spteval$} & \mbox{Description of run-time check} \\\hline
    \sptoptional{\spteval_0} & \mbox{Accepts either the none value or values that match $\spteval_0$} \\
    \sptrawpydict & \mbox{Accepts any Python dictionary} \\
    \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries parameterized by $\spteval_0$ and $\spteval_1$} \\
    %% \sptchkdict{\spteval_0}{\spteval_1} & \mbox{Accepts checked dictionaries with the same key and value types (${\spteval_0, \spteval_1}$)} \\
    \sptdyn & \mbox{No check needed, accepts any value} \\
    \sptobject & \mbox{Accepts any value (except primitives,~\cref{s:c-types})} \\
    \sptclass_0 & \mbox{Accepts instances of class $\sptclass_0$} \\
    \sptnone & \mbox{Accepts the Python none value} \\
    \sptint & \mbox{Accepts integers} \\
    \sptbool & \mbox{Accepts booleans (the integers \code{0} and \code{1})} \\
    \sptfloat & \mbox{Accepts floats}
  \end{array}\)
\end{table}

\subsubsection{Expression Typing, Cast Insertion}

To better show where \SP{} needs to insert casts, we present a selection of typing rules.
Recall that a program declares variables, functions, and classes~(\cref{f:surface-types}).
These statements come with types and fill an environment ($\sptenv$).
Relative to the current environment, the expressions within each statement must
satisfy the judgment in~\cref{f:eval-types}

The first two rules are for function application.
A typed function may be applied an argument that matches its domain type,
in which case it computes a value that matches its codomain type.
A dynamically-typed variable may be applied to any input (that matches the top
type $\sptobject$) to yield a dynamically-typed result.
The next rules illustrate reads from Python dictionaries and checked dictionaries.
A Python dict may be deferenced by any value; the read promises a dynamically-typed result.
By contrast, a checked dict requires an index value that matches its type
and promises a typed result.
Next we have two rules for classes:
\trule{F-Set} says that writes to a class field must match the declared field
type; and \trule{M-App} shows that typed methods impose similar constraints as
typed fuctions.

The final two rules, \trule{C-Sub} and \trule{Matr}, depend on auxiliary judgments
for consistent subtyping ($\spcompat$) and materialization ($\spmatr$).
Consistent subtyping relates type $\spteval_0$ to type $\spteval_1$ if they are
related by static subtyping ($\spteval_0 \spsubteq \spteval_1$) or if
$\spteval_1$ is the dynamic type.
When this relation holds, it is safe to upcast a value from type $\spteval_0$
to type $\spteval_1$.
Materialization relates the dynamic type to any non-dynamic type.
Occurrences of the \trule{Matr} rule are downcasts that require a
run-time check.

The name \emph{materialization} and the rule \trule{Matr} are inspired by prior
work~\cite{clps-popl-2019}.
By itself, the judgment is merely an upside-down type precision relation~\cite{g-icfp-2013,sv-dls-2008}.
Combined with the typing rule, however, materialization is a concise way to find
where a well-typed program needs to insert casts to ensure soundness.
For example, suppose that $\spf$ is a function from integers to the dynamic
type and that $\spx$ is a variable with the dynamic type.
An application $\spapp{\spf}{\spx}$ can satisfy type $\sptfloat$ using two materializations:

\medskip
\begin{mathpar}
  \inferrule*[left=Matr]{
    \inferrule*{
      \inferrule*[left=Matr]{
        \sptenv_0 \wtexpr \spx : \sptdyn
        \\
        \sptdyn \spmatr \sptint
      }{
        \sptenv_0 \wtexpr \spx : \sptint
      }
    }{
      \sptenv_0 \wtexpr \spapp{\spf}{\spx} : \sptdyn
    }
    \quad
    \sptdyn \spmatr \sptfloat
  }{
    \sptenv_0 \wtexpr \spapp{\spf}{\spx} : \sptfloat
  }

  \mbox{Where $\sptenv_0 = \sptvardef{\spx}{\sptdyn}, \sptfundef{\spf}{\sptint}{\sptdyn}$}
\end{mathpar}
\medskip
Consequently, this type derivation calls for two casts at run-time.

We end here, with materialization, rather than present a semantics
for the formalization.
After all, the main benefit of a full formal semantics is to validate
the behavior of complex expressions---a job
better left to the mechanized Redex model.

Furthermore, our casts-via-materialization rule is more aggressive
than what \SP{} actually does to insert casts because
\SP{} supports interactions with plain Python code.
Because it chooses not to recompile Python modules, \SP{} inserts
casts in an overapproximate way:
\begin{itemize}
  \item
    typed functions and methods begin by checking all their arguments;
  \item
    typed classes check all field writes;
  \item
    checked dictionaries check all writes; and
  \item
    typed code checks the results of function calls, references, etc. whenever there is a materialization from the dynamic type.
\end{itemize}
\Cref{s:impl} explains how the type-directed optimizer avoids some
of these checks.

\begin{figure}[t]
  \fbox{$\sptenv \wtexpr \spexpr : \spteval$}~selected rules
  \begin{mathpar}
    \inferrule*[left=S-App]{
      \sptfundef{\spf_0}{\spteval_0}{\spteval_1} \in \sptenv_0
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
    }{
      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \spteval_1
    }

    \inferrule*[left=D-App]{
      \sptvardef{\spf_0}{\sptdyn} \in \sptenv_0
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
    }{
      \sptenv_0 \wtexpr \spapp{\spf_0}{\spexpr_1} : \sptdyn
    }

    \inferrule*[left=PD-Ref]{
      \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \sptobject
    }{
      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \sptdyn
    }

    \inferrule*[left=CD-Ref]{
      \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
      \\\\
      \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
    }{
      \sptenv_0 \wtexpr \spdictref{\spexpr_0}{\spexpr_1} : \spteval_1
    }

    %% TODO show dict write rules?
    %\inferrule*[left=PD-Set]{
    %  \sptenv_0 \wtexpr \spexpr_0 : \sptrawpydict
    %  \\\\
    %  \sptenv_0 \wtexpr \spexpr_1 : \sptobject
    %  \\\\
    %  \sptenv_0 \wtexpr \spexpr_2 : \sptobject
    %}{
    %  \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1}{\spexpr_2} : \sptnone
    %}
    %
    %\inferrule*[left=CD-Set]{
    %  \sptenv_0 \wtexpr \spexpr_0 : \sptchkdict{\spteval_0}{\spteval_1}
    %  \\\\
    %  \sptenv_0 \wtexpr \spexpr_1 : \spteval_0
    %  \\\\
    %  \sptenv_0 \wtexpr \spexpr_2 : \spteval_1
    %}{
    %  \sptenv_0 \wtexpr \spdictset{\spexpr_0}{\spexpr_1} : \sptnone
    %}

    %\inferrule*[left=F-Ref]{
    %  \spann{\spx_0}{\spteval_0} \in \spenvapp{\sptenv_0}{\spc_0}
    %}{
    %  \sptenv_0 \wtexpr \spobjref{\spc_0}{\spx_0} : \spteval_0
    %}
